---
title: "Hypothesis Testing"
subtitle: "Master Seminar Quantitative Research Methods"
author: Peter Putz
date: Institut für Organisation und Globale Managementstudien, JKU Linz
output: 
  ioslides_presentation:
    widescreen: true
    logo: ../../assets/img/JKUWappen.png
    css: ../../assets/css/jku.css
    
---

<style>
table.rmdtable th {
  color: white;
  font-size: 18px;
  background: -webkit-gradient(linear, 50% 0%, 50% 100%, color-stop(40%, #1e786d), color-stop(80%, #144f48)) no-repeat;
  background: -webkit-linear-gradient(top, #1e786d 40%, #144f48 80%) no-repeat;
  background: -moz-linear-gradient(top, #1e786d 40%, #144f48 80%) no-repeat;
  background: -o-linear-gradient(top, #1e786d 40%, #144f48 80%) no-repeat;
  background: linear-gradient(top, #1e786d 40%, #144f48 80%) no-repeat;
}
</style>


## Hypothesis testing

- Hypothesis testing is concerned with **making decisions** using data
- A null hypothesis is specified that represents the **status quo**,
  usually labeled **$H_0$**
- The null hypothesis is assumed true and **statistical evidence is required
  to reject** it in favor of a research or alternative hypothesis 


# Test Strategy


## Example

- In an industry benchmark study a customer satisfaction index (CSI) of more than 30 is 
  considered evidence of excellent customer satisfaction.
- Suppose that in a sample of 100 customers the CSI was 32 with a standard deviation of 10.
- We might want to test the hypothesis that 

    * $H_0 : \mu = 30$
    * $H_a : \mu > 30$
    * where $\mu$ is the population mean CSI.
  

## Test Strategy

- A reasonable strategy would reject the null hypothesis if  
  $\bar x$ was larger than some constant, say $C$
* Typically, $C$ is chosen so that the probability of $\alpha$ is $0.05$   
  (or some other relevant constant)
* $\alpha$ = Type I error rate: Probability of rejecting the null hypothesis when,  
  in fact, the null hypothesis is correct

## Test Strategy

$\bar{x} > C$

```{r, echo=FALSE, fig.width=6, fig.height=4, fig.align='center'}

## Code adapted from Kristoffer Magnusson
## http://rpsychologist.com/creating-a-typical-textbook-illustration-of-statistical-power-using-either-ggplot-or-base-graphics

library(ggplot2)
library(grid)  # need for arrow()
m1 <- 0        # mu H0
sd1 <- 1.5     # sigma H0
m2 <- 3.5      # mu HA
sd2 <- 1.5     # sigma HA

z_crit <- qnorm(1-(0.05/2), m1, sd1)

# set length of tails
min1 <- m1-sd1*4
max1 <- m1+sd1*4
min2 <- m2-sd2*4
max2 <- m2+sd2*4          


x <- seq(min(min1,min2), max(max1, max2), .01) # create x sequence
y1 <- dnorm(x, m1, sd1)                        # normal dist #1
y2 <- dnorm(x, m2, sd2)                        # normal dist #2

df1 <- data.frame("x" = x, "y" = y1)
df2 <- data.frame("x" = x, "y" = y2)

# Alpha polygon
y.poly <- pmin(y1,y2)
poly1 <- data.frame(x=x, y=y.poly)
poly1 <- poly1[poly1$x >= z_crit, ] 
poly1<-rbind(poly1, c(z_crit, 0))  # add lower-left corner

# Beta polygon
poly2 <- df2
poly2 <- poly2[poly2$x <= z_crit,] 
poly2<-rbind(poly2, c(z_crit, 0))  # add lower-left corner

# power polygon; 1-beta
poly3 <- df2
poly3 <- poly3[poly3$x >= z_crit,] 
poly3 <-rbind(poly3, c(z_crit, 0))  # add lower-left corner

# combine polygons. 
poly1$id <- 3 # alpha, give it the highest number to make it the top layer
poly2$id <- 2 # beta
poly3$id <- 1 # power; 1 - beta
poly <- rbind(poly1, poly2, poly3)
poly$id <- factor(poly$id,  labels=c("power","beta","alpha"))


# plot with ggplot2
ggplot(poly[poly$id=="alpha", ], aes(x,y, fill=id, group=id)) +
  geom_line(data=df1, aes(x,y, color="H0", group=NULL, fill=NULL), size=1.5, show_guide=F) + 
  geom_segment(aes(x = 0, y = .02, xend = 0, yend = .3), size=1, linetype="dotted") +
  geom_segment(aes(x = 3.5, y = 0.02, xend = 3.5, yend = .3), size=1, linetype="dotted") +
  
  scale_color_manual("Group", values= c("HA" = "#981e0b","H0" = "black")) +

  annotate("text", label="mu == 30", x=0, y=0, parse=T, size=8) +
  annotate("text", label="bar(x) == 32", x=5.5, y=.15, parse=T, size=8) +

  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.title = element_text(size=22))
```


## Test Strategy

$\bar{x} > C$

```{r, echo=FALSE, fig.width=6, fig.height=4, fig.align='center'}

ggplot(poly[poly$id=="alpha", ], aes(x,y, fill=id, group=id)) +
  geom_polygon(show_guide=F, alpha=I(8/10)) +
  geom_line(data=df1, aes(x,y, color="H0", group=NULL, fill=NULL), size=1.5, show_guide=F) + 
  geom_vline(xintercept = z_crit, size=1, linetype="dashed", color="red") +
  geom_segment(aes(x = 0, y = .02, xend = 0, yend = .3), size=1, linetype="dotted") +
  geom_segment(aes(x = 3.5, y = 0.02, xend = 3.5, yend = .3), size=1, linetype="dotted") +
  
  scale_color_manual("Group", values= c("HA" = "#981e0b","H0" = "black")) +

  annotate("segment", x=4, y=0.043, xend=3.4, yend=0.01, 
           arrow = arrow(length = unit(0.3, "cm")), size=1, color="red") +
  annotate("segment", x=3.2, y=0.27, xend=4.7, yend=0.27, 
           arrow = arrow(length = unit(0.3, "cm")), size=.8, color="red") +
  annotate("segment", x=2.6, y=0.27, xend=1.1, yend=0.27, 
           arrow = arrow(length = unit(0.3, "cm")), size=.8, color="red") +
  annotate("text", label="alpha", x=4.2, y=0.05, parse=T, size=8, color="red") +
  annotate("text", label="mu == 30", x=0, y=0, parse=T, size=8) +
  annotate("text", label="italic(C)", x=2, y=.29, parse=T, size=8, color="red") +
  annotate("text", label="bar(x) == 32", x=5.5, y=.15, parse=T, size=8) +

  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.title = element_text(size=22))
```




## Example continued

- Suppose we choose $\alpha = 0.05$:

    $Z_{1-\alpha} = Z_{0.95} = 1.645$
    
- Hence, $C = 30 + 1.645 = 31.645$
- Since our mean $\bar{x} = 32$ we reject the null hypothesis

\
```{r}
qnorm(.95)
```






## Discussion
* In general we don't convert $C$ back to the original scale
* We would just reject because the Z-score; which is how many
  standard errors the sample mean is above the hypothesized mean
  $$
  \frac{32 - 30}{10 / \sqrt{100}} = 2
  $$
  is greater than $1.645$
* Or, whenever $\sqrt{n} (\bar X - \mu_0) / s > Z_{1-\alpha}$


## General rules
- The $Z$ test for $H_0:\mu = \mu_0$ versus 
    - $H_1: \mu < \mu_0$
    - $H_2: \mu \neq \mu_0$
    - $H_3: \mu > \mu_0$ 
- Test statistic $TS = \frac{\bar{X} - \mu_0}{S / \sqrt{n}}$
- Reject the null hypothesis when 
    - $TS \leq -Z_{1 - \alpha}$
    - $|TS| \geq Z_{1 - \alpha / 2}$
    - $TS \geq Z_{1 - \alpha}$


## Hypothesis testing
* The alternative hypotheses are typically of the form $<$, $>$ or $\neq$
* Note that there are four possible outcomes of our statistical decision process

Truth | Decide | Result |
---|---|---|
$H_0$ | $H_0$ | Correctly accept null |
$H_0$ | $H_a$ | Type I error |
$H_a$ | $H_a$ | Correctly reject null |
$H_a$ | $H_0$ | Type II error |


## Notes
* We have fixed $\alpha$ to be low, so if we reject $H_0$ (either
  our model is wrong) or there is a low probability that we have made
  an error
* We have not fixed the probability of a type II error, $\beta$;
  therefore we tend to say "Fail to reject $H_0$"" rather than
  accepting $H_0$
* Statistical significance is no the same as scientific
  significance


## Small Sample Sizes
* The $Z$ test requires the assumptions of the CLT and for $n$ to be large enough
  for it to apply
* If $n$ is small, then a Gossett's $T$ test is performed exactly in the same way,
  with the normal quantiles replaced by the appropriate Student's $T$ quantiles and
  $n-1$ df



## Normal vs. T Distribution

```{r, echo=FALSE, fig.align='center', fig.width=8, fig.height=5}
x <- seq(-5, 5, by=.01)
y1 <- dnorm(x, 0, 1)
y2  <- dt(x, 2)
y3  <- dt(x, 10)

label1 <- rep("normal", length(x))
label2 <- rep("df = 2", length(x))
label3 <- rep("df = 10", length(x))


df1 <- data.frame("x" = x, "y" = y1, "label" = label1)
df2 <- data.frame("x" = x, "y" = y2, "label" = label2)
df3 <- data.frame("x" = x, "y" = y3, "label" = label3)
df  <- rbind(df1, df2, df3)


ggplot(df, aes(x,y, color=label, group=label)) +
  geom_line(size=1) + 
  scale_colour_discrete(guide = guide_legend(title = NULL)) +
  labs(x="", y="") +
  theme_bw(base_size=20)+
  theme(legend.position = c(0.85, 0.8),
        legend.key = element_blank())
```


## Example Reconsidered (Small Sample Size)
- Consider our example again. Suppose that $n= 16$ (rather than
$100$). Then consider that
$$
.05 = P\left(\frac{\bar X - 30}{s / \sqrt{16}} \geq t_{1-\alpha, 15} ~|~ \mu = 30 \right)
$$
- So that our test statistic is now $\sqrt{16}(32 - 30) / 10 = 0.8$, while the critical value is $t_{1-\alpha, 15} = 1.75$. 
- We now fail to reject.

```{r}
qt(.95, df=15)
```



## Two sided tests {.smaller}
* Suppose that we would reject the null hypothesis if in fact the 
  mean was too large or too small
* That is, we want to test the alternative $H_a : \mu \neq 30$
  (doesn't make a lot of sense in our setting)
* Then note
$$
 \alpha = P\left(\left. \left|\frac{\bar X - 30}{s /\sqrt{16}}\right| > t_{1-\alpha/2,15} ~\right|~ \mu = 30\right)
$$
* That is we will reject if the test statistic, $0.8$, is either
  too large or too small, but the critical value is calculated using
  $\alpha / 2$
* In our example the critical value is $2.13$, so we fail to reject.

```{r}
qt((1-0.05/2), 15)
```



# t-tests in R

## Example 1: Father vs. Son Heights

```{r, message=FALSE}
library(UsingR); data(father.son)
head(father.son, 5)
```


## Example 1 cont'd {.smaller}

Compare means of two paired variables with **one sample t-test**.

$H_0$: Difference between father and son heights $= 0$.  
$H_a$: Diffenence in means $\neq 0$.

```{r}
t.test(father.son$fheight - father.son$sheight)
```



## Example 1 cont'd {.smaller}

Same test with **two sample paired t-test**

```{r}
library(tidyr); fs <- gather(father.son, person, height)
head(fs, 3)
t.test(height ~ person, data = fs, paired = TRUE, alternative = "two.sided")
```


## Example 2 {.smaller}

**Two sample t test** (not paired)

```{r}
head(iris, 3)
sv <- iris[iris$Species %in% c("setosa", "versicolor"),]
t.test(Sepal.Length ~ Species, data = sv, conf.level = 0.99)
```


## Example 3 {.smaller}

**ANOVA**

$H_0: \mu_{setosa} = \mu_{versicolor} = \mu_{virginia}$  
$H_a:$ not all $\mu_i$ ($i =$ setosa, versicolor, virginia) are equal

```{r}
lm <- lm(Sepal.Length ~ Species, data = iris)
anova(lm)
```



## P-values
* The P-value is the probability under the null hypothesis of obtaining evidence as extreme or more extreme than would be observed by chance alone
* If the P-value is small, then either $H_0$ is true and we have observed a rare event or $H_0$ is false
*  In our example the $T$ statistic was $0.8$. 
  * What's the probability of getting a $T$ statistic as large as $0.8$?
```{r}
pt(0.8, 15, lower.tail = FALSE) 
```
* Therefore, the probability of seeing evidence as extreme or more extreme than that actually obtained under $H_0$ is `r pt(0.8, 15, lower.tail = FALSE)`




## Notes
* By reporting a P-value the reader can perform the hypothesis
  test at whatever $\alpha$ level he or she chooses
* If the P-value is less than $\alpha$ you reject the null hypothesis 
* For two sided hypothesis test, double the smaller of the two one
  sided hypothesis test P-values


# Power of Hypothesis Test



## Power
- Power is the probability of rejecting the null hypothesis when it is false
- Ergo, power (as it's name would suggest) is a good thing; you want more power
- A type II error (a bad thing, as its name would suggest) is failing to reject the null hypothesis when it's false; the probability of a type II error is usually called $\beta$
- Note Power  $= 1 - \beta$


## Power

```{r, echo=FALSE, fig.align='center', fig.width=7, fig,highlight=4}
m1 <- 0  # mu H0
sd1 <- 1.5 # sigma H0
m2 <- 3.5 # mu HA
sd2 <- 1.5 # sigma HA

z_crit <- qnorm(1-(0.05/2), m1, sd1)

# set length of tails
min1 <- m1-sd1*4
max1 <- m1+sd1*4
min2 <- m2-sd2*4
max2 <- m2+sd2*4          
# create x sequence
x <- seq(min(min1,min2), max(max1, max2), .01)
# generate normal dist #1
y1 <- dnorm(x, m1, sd1)
# put in data frame
df1 <- data.frame("x" = x, "y" = y1)
# generate normal dist #2
y2 <- dnorm(x, m2, sd2)
# put in data frame
df2 <- data.frame("x" = x, "y" = y2)

# Alpha polygon
y.poly <- pmin(y1,y2)
poly1 <- data.frame(x=x, y=y.poly)
poly1 <- poly1[poly1$x >= z_crit, ] 
poly1<-rbind(poly1, c(z_crit, 0))  # add lower-left corner

# Beta polygon
poly2 <- df2
poly2 <- poly2[poly2$x <= z_crit,] 
poly2<-rbind(poly2, c(z_crit, 0))  # add lower-left corner

# power polygon; 1-beta
poly3 <- df2
poly3 <- poly3[poly3$x >= z_crit,] 
poly3 <-rbind(poly3, c(z_crit, 0))  # add lower-left corner

# combine polygons. 
poly1$id <- 3 # alpha, give it the highest number to make it the top layer
poly2$id <- 2 # beta
poly3$id <- 1 # power; 1 - beta
poly <- rbind(poly1, poly2, poly3)
poly$id <- factor(poly$id,  labels=c("power","beta","alpha"))




# plot with ggplot2
ggplot(poly, aes(x,y, fill=id, group=id)) +
  geom_polygon(show_guide=F, alpha=I(8/10)) +
  # add line for treatment group
  geom_line(data=df1, aes(x,y, color="H0", group=NULL, fill=NULL), size=1.5, show_guide=F) + 
  # add line for treatment group. These lines could be combined into one dataframe.
  geom_line(data=df2, aes(color="HA", group=NULL, fill=NULL),size=1.5, show_guide=F) +
  # add vlines for z_crit
  geom_vline(xintercept = z_crit, size=1, linetype="dashed") +
  # change colors 
  scale_color_manual("Group", 
                     values= c("HA" = "#981e0b","H0" = "black")) +
  scale_fill_manual("test", values= c("alpha" = "#0d6374","beta" = "#be805e","power"="#7cecee")) +
  # beta arrow
  annotate("segment", x=0.1, y=0.045, xend=1.3, yend=0.01, 
           arrow = arrow(length = unit(0.3, "cm")), size=1) +
  annotate("text", label="beta", x=0, y=0.05, parse=T, size=8) +
  # alpha arrow
  annotate("segment", x=4, y=0.043, xend=3.4, yend=0.01, 
           arrow = arrow(length = unit(0.3, "cm")), size=1) +
  annotate("text", label="frac(alpha,2)", x=4.2, y=0.05, parse=T, size=8) +
  # power arrow
  annotate("segment", x=6, y=0.2, xend=4.5, yend=0.15, 
           arrow = arrow(length = unit(0.3, "cm")), size=1) +
  annotate("text", label="1-beta ~~ (power)", x=7.3, y=0.22, parse=T, size=8) +
  # H_0 title
  annotate("text", label="H[0]", x=m1, y=0.28, parse=T, size=8) +
  # H_a title
  annotate("text", label="H[a]", x=m2+0.3, y=0.28, parse=T, size=8) +
# ggtitle("Statistical Power Plots, Textbook-style") +
  # remove some elements
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.background = element_blank(),
     #  plot.background = element_rect(fill="#f9f0ea"),
        panel.border = element_blank(),
        axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        plot.title = element_text(size=22))
```



## Notes
- Consider our previous example involving CSI
- $H_0: \mu = 30$ versus $H_a: \mu > 30$
- Then power is 
$$P\left(\frac{\bar X - 30}{s /\sqrt{n}} > t_{1-\alpha,n-1} ~|~ \mu = \mu_a \right)$$
- Note that this is a function that depends on the specific value of $\mu_a$!
- Notice as $\mu_a$ approaches $30$ the power approaches $\alpha$



## Calculating Power for Gaussian Data
Assume that $n$ is large and that we know $\sigma$
$$
\begin{align}
1 -\beta & = 
P\left(\frac{\bar X - 30}{\sigma /\sqrt{n}} > z_{1-\alpha} ~|~ \mu = \mu_a \right)\\
& = P\left(\frac{\bar X - \mu_a + \mu_a - 30}{\sigma /\sqrt{n}} > z_{1-\alpha} ~|~ \mu = \mu_a \right)\\ \\
& = P\left(\frac{\bar X - \mu_a}{\sigma /\sqrt{n}} > z_{1-\alpha} - \frac{\mu_a - 30}{\sigma /\sqrt{n}} ~|~ \mu = \mu_a \right)\\ \\
& = P\left(Z > z_{1-\alpha} - \frac{\mu_a - 30}{\sigma /\sqrt{n}} ~|~ \mu = \mu_a \right)\\ \\
\end{align}
$$


## Power for CSI Example

$$
\begin{align}
1 -\beta & = P\left(Z > z_{1-\alpha} - \frac{\mu_a - 30}{\sigma /\sqrt{n}}\right)\\
         & = P\left(Z > z_{0.95} - \frac{32 - 30}{10 /\sqrt{100}}\right)\\
         & = P\left(Z > 1.645 - 2\right)\\
         & = P(Z > -0.355)\\
         & = .639
\end{align}
$$

```{r}
z  <- qnorm(0.95)
pnorm(z - 2, lower.tail = FALSE)
```


## Calculating T-Test Power with R

```{r}
power.t.test(power = NULL, n = 100, delta = 2, sd = 10, 
             type = "one.sample",  alt = "one.sided")
```


## Finding Minimum Sample Size

Suppose we want a power of 80%, what is the minimum sample size?

```{r}
power.t.test(power = 0.8, n = NULL, delta = 2, sd = 10, 
             type = "one.sample",  alt = "one.sided")$n
```

power.t.test automatically solves the equation for the parameter set to NULL.



## Acknowledgements

\
\
\

<div class="centered">
This class and many slides were inspired by:

**Brian Caffo, Jeff Leek, and Roger Peng**   
Department of Biostatistics   
Johns Hopkins Bloomberg School of Public Health   

and the  

[Coursera Data Science Specialization Program](https://www.coursera.org/specialization/jhudatascience/1?utm_medium=courseDescripTop)

</div>

