---
title: "Number of Scientists in Rich and Poor Countries"
author: "Peter Putz"
date: "May 10, 2015"
output:
  html_document:
    highlight: pygments
    theme: united
---


## Download Data

1. Download .zip file and extract data files.
2. Store paths to data files in vector "files".

```{r download}

if(!file.exists("./data")){
#  dir.create("./data")
  url <- "http://api.worldbank.org/v2/en/topic/14?downloadformat=csv"
  download.file(url,destfile="./temp.zip")
  unzip("./temp.zip", exdir = "./data")
  file.remove("./temp.zip")
  }

files <- paste("./data", list.files("./data"), sep="/")
files
```


## Read Data into R

1. Read main data file (14\_Topic\_en\_csv\_v2.csv). This includes defining the column types.
2. Read the two metadata files.

```{r readdata, message=FALSE}

library(readr)
library(dplyr)
library(tidyr)
library(knitr)

col_types = paste("cccc", paste(rep("d", 55), collapse = ""), "_", sep="")
RND <- read_csv(files[2], skip=2, col_types = col_types)
#glimpse(RND)

RND.C <- read_csv(files[3])
RND.I <- read_csv(files[4])
```

## Number of Scientists in Rich and Poor Countries

#### Step 1: Data Wrangling

1. Subset data ("Researchers in R&D (per million people)") and calculate average for the years 2009 thru 2013.
2. Add variables "Region" and "IncomeGroup" from metadata file RND.C


```{r scientists}
RND.Staff  <- RND %>%
  filter(grepl("SP.POP.S", `Indicator Code`)) %>%
  gather(key=Year, Researchers, `1960`:`2014`, na.rm = T, convert=T) %>%
  filter(Year %in% 2009:2013) %>%
  group_by(`Country Name`, `Country Code`) %>%
  summarize(Scientists = mean(Researchers))
  
```

```{r join}
RND.Staff1 <- RND.Staff %>%
  left_join(RND.C[,2:4], by = "Country Code") %>%
  filter(IncomeGroup != "")
```

#### Step 2: Analysis

1. Create table with income groups in rows and summary stats in columns.
2. Create boxplot; this includes sorting the factor levels for "IncomeGroup".

```{r sumtable, results='asis'}
RND.Staff2  <- RND.Staff1 %>%
  group_by(IncomeGroup) %>%
  summarize(`Number Of Countries` = n(),
            Minimum = min(Scientists),
            `1st Quantile` = quantile(Scientists, 0.25),
            Median = median(Scientists),
            `3rd Quantile` = quantile(Scientists, 0.75),
            Maximum = max(Scientists))

kable(RND.Staff2)
```


```{r boxplot}
library(ggplot2)

RND.Staff1$IncomeGroup <- factor(RND.Staff1$IncomeGroup,
                                 levels=c("High income: OECD",
                                         "High income: nonOECD",
                                         "Upper middle income",
                                         "Lower middle income",
                                         "Low income"))

bp  <- ggplot(RND.Staff1, aes(IncomeGroup, Scientists)) + 
              geom_boxplot() +
              coord_flip() +
              scale_x_discrete(limits = rev(levels(RND.Staff1$IncomeGrou))) +
              labs(title = "Number of Researchers per One Million People",
                   x="", y="")
bp
```


## Interactive Map of Euope and Central Asia

1. Subset data for "Europe & Central Asia".
2. Create `googleVis` chart.

```{r europe}
RND.Eu  <- RND.Staff1 %>%
  filter(Region == "Europe & Central Asia") %>%
  rename(Country = `Country Name`) %>%
  select(Country, Scientists)
```


```{r googlevis, results='asis'}
suppressPackageStartupMessages(library(googleVis))

map_1<- gvisGeoChart(RND.Eu, locationvar='Country', 
                     colorvar='Scientists',
                     options=list(
                       projection="kavrayskiy-vii",
                       region="150",
                       colors="['#E6E6F5', '#00006B']",
                       width=900, height=600
                     ))


print(map_1, tag="chart")
```


## Session Info


```{r session}
sessionInfo()
```
