---
title: "Manipulating Data"
subtitle: Master Seminar Quantitative Research Methods
author: "Peter Putz"
date: "Institut für Organisation und Globale Managementstudien, JKU Linz"
output:
  ioslides_presentation:
    css: ../../assets/css/jku.css
    highlight: pygments
    logo: ../../assets/img/JKUWappen.png
    widescreen: yes
---

<style>
table.rmdtable th {
  color: white;
  font-size: 18px;
  background: -webkit-gradient(linear, 50% 0%, 50% 100%, color-stop(40%, #1e786d), color-stop(80%, #144f48)) no-repeat;
  background: -webkit-linear-gradient(top, #1e786d 40%, #144f48 80%) no-repeat;
  background: -moz-linear-gradient(top, #1e786d 40%, #144f48 80%) no-repeat;
  background: -o-linear-gradient(top, #1e786d 40%, #144f48 80%) no-repeat;
  background: linear-gradient(top, #1e786d 40%, #144f48 80%) no-repeat;
}
</style>


## Manipulating Data 


How much time is typically spent in wrestling with the data and "beating it into shape" so that the actual analytics can begin?

**The most common response is 80%.**


[Dan Steinberg](http://info.salford-systems.com/blog/bid/299181/How-Much-Time-Needs-to-be-Spent-Preparing-Data-for-Analysis)



## Recommended R Packages

- `tidyr`: Get raw data ready for analysis
- `dplyr`: Data manipulation (subset, group, etc.)


# tidyr

## Tidy Data

The two most important properties of tidy data are:

- Each column is a variable.
- Each row is an observation.

## tidyr

tidyr provides three main functions for tidying your messy data: 

- `gather()` 
- `spread()`
- `separate()`


## `gather()` {.smaller}

Takes multiple columns, and gathers them into key-value pairs: it makes "wide"" data longer.

The following data set represents four insurance sales reps and the number of car and home policies they sold in different regions.

```{r, message=FALSE, echo=FALSE}
library(tidyr); library(dplyr)
set.seed(10)
messy <- data.frame(
  id = c("Simon", "Rojeen", "Alia", "Sunny"),
  car.RO = sample(20:120, 4, replace=T),
  home.RO = sample(20:120, 4, replace=T),
  car.UU = sample(20:120, 4, replace=T),
  home.UU = sample(20:120, 4, replace=T),
  car.FR = sample(20:120, 4, replace=T),
  home.FR = sample(20:120, 4, replace=T)
)


print(messy)
```



## `gather()`

We have three variables (region, policy and sales), but only region is currently in a column. 

We use `gather()` to gather the car and rental columns into key-value pairs of policy and sales:

```{r}
tidier <- gather(data = messy, key = key, value = sales, 
         car.RO:home.FR)
print(tidier)
```



## `spread()`

`spread()` is the reserve of `gather()`. It takes two columns (a key-value pair) and spreads them in to multiple columns, making “long” data wider.

```{r}
spread(data = tidier, key = key, value = sales)
```



## `separate()`

Next we use separate() to split the key into location and time

```{r}
tidy  <- separate(data = tidier, col = key, into = c("policy", "region"))
head(tidy, 10)
```


# dplyr

## dplyr Verbs

-------------  -----------------------------
`filter()`     subset rows
`arrange()`    sort rows
`select()`     subset colums
`distinct()`   get unique rows
`mutate()`     calculate new column
`summarise()`  calclute summary stats (mean, max, ...)
`sample_n()`   subset random sample
-------------  -----------------------------


## `filter()`

Subset data for regions RO and UU

```{r}
filter(tidy, region == "RO" | region == "UU")
```

## Relational and Logical Opertors {.smaller}

Operator | Use                     | Operator    | Use 
------- | ------------------------ | ----------- | ------------------
x < y   | less than                | &           | AND
x > y   | greater than             | \|          | OR
x <= y  | less or equal than       | %in%        | match
x >= y  | greater of equal than    | grepl       | pattern matching
x == y  | equal                    | \           | \   
x != y  | not equal                | \           | \   

\

**The output is always a logical vector (TRUE, FALSE, NA)**



## `filter()`

```r
filter(tidy, region == "RO" | region == "UU")
filter(tidy, region %in% c("UU", "RO"))
filter(tidy, region != "FR")
```
The above statements yield the same results.




## `arrange()`

Sort by id and policy.

```{r}
arrange(tidy, id, policy)
```




## `select()`, `distinct()`

Get a list of staff and the kind of policies sold.

```{r}
distinct(select(tidy, id, policy))
```



## `summarize()`

Calculate sums over all regions.

```{r}
res  <-  summarize(group_by(tidy, id, policy), subtotal = sum(sales))
res
```



## Piping with `%>%`

```r
summarize(group_by(tidy, id, policy), subtotal = sum(sales))
```
equals 

```r
tidy %>%
  group_by(id, policy) %>%
  summarize(subtotal = sum(sales))
```


## Calculating Margin Percentages 

```{r}
res %>%
  group_by(policy) %>%
  arrange(policy) %>%
  mutate(total = sum(subtotal), 
         percent = subtotal / sum(subtotal) * 100)
```


## Resources

- [Data Wrangling Cheat Sheet](http://www.rstudio.com/resources/cheatsheets/)
