---
title: "Reading and Writing Data With R"
subtitle: Master Seminar Quantitative Research Methods
author: "Peter Putz"
date: "Institut für Organisation und Globale Managementstudien, JKU Linz"
output:
  ioslides_presentation:
    css: ../../assets/css/jku.css
    highlight: pygments
    logo: ../../assets/img/JKUWappen.png
    widescreen: yes

---

<style>
table.rmdtable th {
  color: white;
  font-size: 18px;
  background: -webkit-gradient(linear, 50% 0%, 50% 100%, color-stop(40%, #1e786d), color-stop(80%, #144f48)) no-repeat;
  background: -webkit-linear-gradient(top, #1e786d 40%, #144f48 80%) no-repeat;
  background: -moz-linear-gradient(top, #1e786d 40%, #144f48 80%) no-repeat;
  background: -o-linear-gradient(top, #1e786d 40%, #144f48 80%) no-repeat;
  background: linear-gradient(top, #1e786d 40%, #144f48 80%) no-repeat;
}
</style>

# Reading, Writing Data with Base R


## Reading Data

There are a few principal functions reading data into R.

------------------------ | ---------------------------------------------
`read.table`, `read.csv` | reading tabular data
`readLines`              | reading lines of a text file
`load`                   | reading in saved workspaces
`source`                 | reading in R code files (`inverse` of `dump`) 
`dget`                   | reading in R code files (`inverse` of `dput`)




## Writing Data

There are analogous functions for writing data to files

- write.table
- writeLines
- save
- dump
- dput


## Sidenote: Get/set your working directory

- A basic component of working with data is knowing your working directory
- The two main commands are ```getwd()``` and ```setwd()```. 
- Be aware of relative versus absolute paths
    - __Relative__: ```setwd("./data")```, ```setwd("../")```
    - __Absolute__: ```setwd("/Users/pp/Desktop/data/")```
- Important difference in Windows ```setwd("C:\\Users\\Andrew\\Downloads")``` or ```setwd("C:/Users/Andrew/Desktop/data/")```
- [RStudio and working directories](https://support.rstudio.com/hc/en-us/articles/200711843-Working-Directories-and-Workspaces)



## Reading Data Files with read.table {.smaller}

For small to moderately sized datasets, you can usually call read.table without specifying any other arguments

```r
setwd("/Users/pp/Desktop/samples/")
getwd()
data <- read.table("co2_mm_gl.txt")
```

R will automatically

- skip lines that begin with a #
- figure out how many rows there are (and how much memory needs to be allocated)
- figure what type of variable is in each column of the table

Telling R all these things directly makes R run faster and more efficiently.

- `read.csv` is identical to read.table except that the default separator is a comma.



## read.table

If needed, use the following arguments:

- `file`, the name of a file, or a connection
- `header`, logical indicating if the file has a header line
- `sep`, a string indicating how the columns are separated
- `colClasses`, a character vector indicating the class of each column in the dataset 
- `nrows`, the number of rows in the dataset
- `comment.char`, a character string indicating the comment character
- `skip`, the number of lines to skip from the beginning
- `stringsAsFactors`, should character variables be coded as factors?



## Reading in Larger Datasets with read.table

With much larger datasets, doing the following things will make your life easier and will prevent R from choking.

- Read the help page for read.table, which contains many hints
- Make a rough calculation of the memory required to store your dataset. If the dataset is larger than the amount of RAM on your computer, you can probably stop right here.
- Set `comment.char = ""` if there are no commented lines in your file.

\
**Alternative**: Use package `readr`


<!--


## Textual Formats

- Dumpin and dputing are useful because the resulting textual format is edit-able, and in the case of corruption, potentially recoverable.
- Unlike writing out a table or csv file, `dump` and `dput` preserve the _metadata_ (sacrificing some readability), so that another user doesn’t have to specify it all over again.
- Textual formats can work much better with version control programs like subversion or git which can only track changes meaningfully in text files
- Textual formats can be longer-lived; if there is corruption somewhere in the file, it can be easier to fix the problem
- Textual formats adhere to the “Unix philosophy” 
- Downside: The format is not very space-efficient



## dput-ting R Objects

Another way to pass data around is by deparsing the R object with dput and reading it back in using `dget`.

```r
> y <- data.frame(a = 1, b = "a")
> dput(y)
structure(list(a = 1,
               b = structure(1L, .Label = "a",
                             class = "factor")),
          .Names = c("a", "b"), row.names = c(NA, -1L),
          class = "data.frame")
> dput(y, file = "y.R")
> new.y <- dget("y.R")
> new.y
   a  b 
1  1  a
```



## Dumping R Objects

Multiple objects can be deparsed using the dump function and read back in using `source`.

```r
> x <- "foo"
> y <- data.frame(a = 1, b = "a")
> dump(c("x", "y"), file = "data.R") 
> rm(x, y)
> source("data.R")
> y
  a  b 
1 1  a
> x
[1] "foo"
```



## Interfaces to the Outside World

Data are read in using _connection_ interfaces. Connections can be made to files (most common) or to other more exotic things.

- `file`, opens a connection to a file
- `gzfile`, opens a connection to a file compressed with gzip
- `bzfile`, opens a connection to a file compressed with bzip2 
- `url`, opens a connection to a webpage



## File Connections

```r
> str(file)
function (description = "", open = "", blocking = TRUE,
          encoding = getOption("encoding"))
```

- `description` is the name of the file 
- `open` is a code indicating
    - “r” read only
    - “w” writing (and initializing a new file)
    - “a” appending
    - “rb”, “wb”, “ab” reading, writing, or appending in binary mode (Windows)



## Connections

In general, connections are powerful tools that let you navigate files or other external objects. In practice, we often don’t need to deal with the connection interface directly.

```r
con <- file("foo.txt", "r")
data <- read.csv(con)
close(con)
```

is the same as

```r
data <- read.csv("foo.txt")
```



## Reading Lines of a Text File

```r
> con <- gzfile("words.gz") 
> x <- readLines(con, 10) 
> x
 [1] "1080"     "10-point" "10th"     "11-point"
 [5] "12-point" "16-point" "18-point" "1st"
 [9] "2"        "20-point"
```

`writeLines` takes a character vector and writes each element one line at a time to a text file.

-->

## Reading Lines of a Text File

`readLines` can be useful for reading in lines of webpages

```{r}
## This might take time
con <- url("http://ti.arc.nasa.gov/profile/pputz/", "r")
x <- readLines(con)
head(x)
```

# Downloading Files From the Internet


## Checking for and creating directories

* ```file.exists("directoryName")``` will check to see if the directory exists
* ```dir.create("directoryName")``` will create a directory if it doesn't exist
* Here is an example checking for a "data" directory and creating it if it doesn't exist

```r
if(!file.exists("data")){
  dir.create("data")
}
```




## Getting data from the internet - download.file()

* Downloads a file from the internet
* Even if you could do this by hand, helps with reproducibility
* Important parameters are _url_, _destfile_, _method_
* Useful for downloading tab-delimited, csv, and other files



## Example - Statistik Austria {.smaller}


Statistic Austria provides sample data for teaching and research:

http://www.statistik.at/web_de/services/mikrodaten_fuer_forschung_und_lehre/kostenloser_download_von_testdaten/index.html



## Example - Income Taxes {.smaller}

```{r taxes}
setwd("~/Desktop")
taxUrl <- "http://www.statistik.at/web_de/static/integrierten_lohn-_und_einkommensteuerstatistik_2011_sds_datensatz_rdata_077823.rdata"
if(!file.exists("data")){
  dir.create("data")
  download.file(taxUrl,destfile="./data/incometaxes.rdata")
}
list.files("./data")
```
```{r load}
load("~/Desktop/data/incometaxes.rdata")
head(SDS_INT_ST2011, 3)
```



## Some notes about download.file()

* If the url starts with _http_ you can use download.file()
* If the url starts with _https_ on Windows you may be ok
* If the url starts with _https_ on Mac you may need to set _method="curl"_
* If the file is big, this might take a while
* Be sure to record when you downloaded. 



#  Packages For Reading Data


## Useful Packages

------------ | ------------------------
readr        | reaplacement for read.table, read.csv
readxl       | Excel files
foreign      | SPSS, SAS, Stata, etc.
XML          | [XML](http://www.w3schools.com/xml/xml_whatis.asp) (E**x**tensible **M**arkup **L**anguage)
jsonlite     | [JSON](http://www.w3schools.com/json/default.asp) (**J**ava**S**cript **O**bject **N**otation)


## readr

`readr` is under active development. It makes it easy to read many types of tabular data:

- Delimited files with `read_delim()`, `read_csv()`, `read_tsv()`, `and read_csv2()`.
- Fixed width files with `read_fwf()`, and `read_table()`.
- Web log files with `read_log()`.


## readr Example {.smaller}

```{r, message=FALSE}
library(readr); library(dplyr)
url <- url("http://www.statistik.at/web_de/static/einbuergerungen-datensatz_2005_forschung_und_lehre_024892.csv")
einwanderungen  <- read_csv2(url)
einwanderungen
```


## readxls {.smaller}

Supports both the legacy .xls format and the modern xml-based .xlsx format.

```{r}
library(readxl)
# Use a excel file included in the package
sample <- system.file("extdata", "datasets.xlsx", package = "readxl")

# Read by tab position
head(read_excel(sample, 2))
```


# Packages for Specific Data Sources

## Packages

------------ | ------------------------
eurostat     | Statistical office of the European Union
Quandl       | Commercial provider of global datasets


## Eurostat Example {.smaller}

```{r eurostatToc, results='asis'}
library(eurostat)

# data sets on employment
kable(head(search_eurostat("employment"), n = 2))
```

\

```{r eurostatDat}
# get data set specified data set
id <- "ei_lmhr_m"
dat <- get_eurostat(id, time_format = "num")
```


## Quandl Example

```{r quandl, warning=FALSE, message=FALSE, eval=FALSE}
library(Quandl)
Quandl.auth("Gxb4ms5yr8hZuqZpQ3DX")
mydata <- Quandl("ODA/USA_NGDP_RPCH", )
```

**Last time I tried, I could not get the package to work correctly.**

However, you can always download the data and use e.g. `readr`.