---
title: "Plotting with ggplot2"
subtitle: Master Seminar Quantitative Research Methods
author: "Peter Putz"
date: "Institut für Organisation und Globale Managementstudien, JKU Linz"
output:
  ioslides_presentation:
    css: ../../assets/css/jku.css
    highlight: pygments
    logo: ../../assets/img/JKUWappen.png
    widescreen: yes

---

<style>
table.rmdtable th {
  color: white;
  font-size: 18px;
  background: -webkit-gradient(linear, 50% 0%, 50% 100%, color-stop(40%, #1e786d), color-stop(80%, #144f48)) no-repeat;
  background: -webkit-linear-gradient(top, #1e786d 40%, #144f48 80%) no-repeat;
  background: -moz-linear-gradient(top, #1e786d 40%, #144f48 80%) no-repeat;
  background: -o-linear-gradient(top, #1e786d 40%, #144f48 80%) no-repeat;
  background: linear-gradient(top, #1e786d 40%, #144f48 80%) no-repeat;
}
</style>



## What is ggplot2?

- An implementation of _The Grammar of Graphics_ by Leland Wilkinson
- Written by Hadley Wickham (while he was a graduate student at Iowa State)
- A “third” graphics system for R (along with __base__ and __lattice__)
- Available from CRAN via `install.packages()`
- Web site: http://ggplot2.org (better documentation)


## What is ggplot2?

- Grammar of graphics represents an abstraction of graphics ideas/objects
- Think “verb”, “noun”, “adjective” for graphics
- Allows for a “theory” of graphics on which to build new graphics and graphics objects
- “Shorten the distance from mind to page”


## Basic Components of a ggplot2 Plot
- A _data frame_
- _aesthetic mappings_: how variables are mapped to axes, color, size 
- _geoms_: geometric objects like points, lines, shapes. 
- _facets_: for conditional plots. 
- _stats_: statistical transformations like binning, quantiles, smoothing. 
- _scales_: what scale an aesthetic map uses (example: male = red, female = blue). 
- _coordinate system_ 



## Example Dataset

```{r}
library(ggplot2)
str(mpg)
```


## Not Quite a Plot Yet | Mapping of data and variables to axes
```{r}
g  <- ggplot(data = mpg, aes(displ, hwy))
summary(g)
```


## Basic Plot with Point Layer

```{r, fig.width=7, fig.height=4}
g  <- ggplot(data = mpg, aes(displ, hwy)) + 
  geom_point()
g
```

## Color-Code Third Variable

```{r, fig.width=7, fig.height=4}
h  <- ggplot(data = mpg, aes(displ, hwy, color = drv)) + 
  geom_point()
h
```

## Adding a Layer: Facets

```{r, fig.width=7, fig.height=4}
g  <- ggplot(data = mpg, aes(displ, hwy)) + 
  geom_point()
g  <- g + facet_grid(facets = . ~ drv)
g
```


## Sidebar: Changing Facet Labels {.smaller}

```{r, fig.width=7, fig.height=4}
levels(mpg$drv) <- c("4-weel-drive", "front-wheel-drive", "rear-wheel-drive")
g  <- ggplot(data = mpg, aes(displ, hwy)) + 
             geom_point() +
             facet_grid(facets = . ~ drv)
g
```

## Adding Another Layer: Regression Lines

```{r, fig.width=7, fig.height=4}
g  <- g + geom_smooth(method = "lm", se = T, color = "red")
g
```

## Annotation
- Labels: `xlab()`, `ylab()`, `labs()`, `ggtitle()`
- Each of the "geom"" functions has options to modify 
- For things that only make sense globally, use `theme()` 
   - Example: `theme(legend.position = "none")` 
- Two standard appearance themes are included
   - `theme_gray()`: The default theme (gray background)
   - `theme_bw()`: More stark/plain 


## Another Layer: Title and Axes Lables {.smaller}

```{r, fig.width=7, fig.height=4}
g <- g + labs(y = "Miles per Gallon (Highway)",
              x = "Displacement in Liters",
              title = "Fuel Efficiency by Type of Drive")
g
```


## Changing Fonts and Backgrounds {.smaller}

```{r, warning=FALSE}
g <- g + theme_bw() + 
  theme(plot.title = element_text(family = "Times", 
                                  color="#666666", 
                                  face="bold", size=24, vjust=2))
g
```


## Final Plot

```{r, fig.width=9, echo=FALSE, warning=FALSE}
levels(mpg$drv) <- c("4-weel-drive", "front-wheel-drive", "rear-wheel-drive")
ggplot(data = mpg, aes(displ, hwy)) + 
  geom_point() +
  facet_grid(facets = . ~ drv) +
  geom_smooth(method = "lm", se = T, color = "red") + 
  labs(y = "Miles per Gallon (Highway)",
       x = "Displacement in Liters",
       title = "Fuel Efficiency by Type of Drive") +
  theme_bw() + 
  theme(plot.title = element_text(family = "Times", color="#666666", 
                                  face="bold", size=32, vjust=2)) 
```

## Code Final Plot

```{r, fig.width=9, eval=FALSE, warning=FALSE}
levels(mpg$drv) <- c("4-weel-drive", "front-wheel-drive", "rear-wheel-drive")
ggplot(data = mpg, aes(displ, hwy)) + 
  geom_point() +
  facet_grid(facets = . ~ drv) +
  geom_smooth(method = "lm", se = T, color = "red") + 
  labs(y = "Miles per Gallon (Highway)",
       x = "Displacement in Liters",
       title = "Fuel Efficiency by Type of Drive") +
  theme_bw() + 
  theme(plot.title = element_text(family = "Times", color="#666666", 
                                  face="bold", size=32, vjust=2)) 
```


## Printing Plots

```{r}
ggsave("/Users/pp/Desktop/fuel.pdf", g, width=6, height=4)
```
Note: File extension drives file format.


## Summary
- ggplot2 is very powerful and flexible if you learn the “grammar” and the various elements that can be tuned/modified
- Many more types of plots can be made; explore and play around with the package 


## Resources

- The _ggplot2_ book by Hadley Wickham
- The _R Graphics Cookbook_ by Winston Chang (examples in base plots and in ggplot2)
- ggplot2 web site (http://ggplot2.org)

