---
title: "Exploration of iris Dataset"
author: "Peter Putz"
date: "April 27, 2015"
output:
  word_document:
    highlight: pygments
  html_document:
    highlight: pygments
---

## Abstract

In this paper we explore the dataset "cars". The data give the speed of cars and the distances taken to stop. Note that the data were recorded in the 1920s.

## Data Set

```{r dim}
dim(cars)
```

The dataset consists of `r dim(cars)[1]` observations and `r dim(cars[2])` variables. The variable names are: `r names(cars)`

```{r head}
head(cars, 5)
```

## Theoretical Breaking Distance

The theoretical breaking distance is determined by the intial speed and the friction coefficient:

$d = \frac{v^2}{2 \mu g}$

d … breaking distance

$\mu$ … friction coeficient

g … gravity of the Earth


Friction Coeficient: $\mu = \frac{v^2}{2dg}$


## Measured Data

```{r distance_plot}
plot(dist ~ speed, data=cars,
     main = "Breaking Distance over Speed")
```

```{r friction}
g  <- 9.81
cars$friction  <- cars$speed^2/(2 * g * cars$dist)
summary(cars$friction)
hist(cars$friction)
plot(friction ~ speed, data=cars)
```

