---
title: "Installing R and RStudio"
subtitle: "Master Seminar Quantitative Research Methods"
author: Peter Putz
date: Institut für Organisation und Globale Managementstudien, JKU Linz
output: 
  ioslides_presentation:
    widescreen: true
    logo: ../../assets/img/jku.svg
    css: ../../assets/css/jku.css
---

## Installing R for Windows

1. Navigate to the (Austrian) [R Project Site](http://cran.at.r-project.org)
2. Click on [Download R for Windows](http://cran.at.r-project.org/bin/windows/) and select "install R for the first time"
3. Download the R for Windows
4. Follow instruction


[Instructions on YouTube](https://youtu.be/hxj0UG4boGU)


## Installing R for Mac

1. Navigate to the (Austrian) [R Project Site](http://cran.at.r-project.org)
2. Click on [Download R for (Mac) OS X](http://cran.at.r-project.org/bin/macosx/) and select "install R for the first time"
3. Download e.g. R-3.2.3.pkg
4. Follow instruction

[Instructions on YouTube](https://youtu.be/_oRuJKIkPEM)



## Installing RStudio

1. Navigate to [RStudio Download Page](http://www.rstudio.com/products/rstudio/download/)
2. Select installer package for your OS. For example:
    - RStudio 0.99.891 - Windows Vista/7/8/10
    - RStudio 0.98.891 - Mac OS X 10.6+ (64-bit)

[Instructions on YouTube](https://youtu.be/bM7Sfz-LADM)



## swirl

`swirl` is an R package that helps you learn "R programming and data science interactively, at your own pace, and right in the R console!"

[http://swirlstats.com](http://swirlstats.com)


## Installing swirl

1) Install the most recent version version of swirl.   
    In R or RStudion enter the following commands:

    ```{r, eval=FALSE}
    install.packages("swirl")
    ```
    
2) Start `swirl` and load a course.  
    Substitute the name of the course that you wish to install:

    ```{r, eval=FALSE}
    library(swirl)
    install_from_swirl("Course Name Here")
    swirl()
    ```

For example, `install_from_swirl("R Programming Alt")` will install the R Programming course.  
**Please note that course names are case sensitive!**


# swirl Demo


## swirl Confrimation

At the end of each lession you have the option notify someone.  
Please send me your confirmation to `peter.putz@jku.at`.

```{r, eval=FALSE}
|============================================================================| 100%

| Would you like to inform someone about your successful completion of this lesson?

1: No
2: Yes

Auswahl: 2
| What is your full name?
Peter Putz
| What is the email address of the person you'd like to notify?'
peter.putz@jku.at
```

    
# R Support

## Built-in Help Functions

>- Use `?`, `??`, or `help()` in the R console
>- Click a package name in RStudio.  
    Some packages have nice vignettes and user guides.


## R Support on the Web

>- [stackoverflow](http://stackoverflow.com/tags/r/info)
>- [R Bloggers](http://www.r-bloggers.com)
>- Google is your friend!


